### 2017-12-08 paper.tex
Initial commit

\documentclass[twocolumn]{article}

\usepackage{minted}
\usepackage{graphicx}
\usepackage{titlesec}
\usepackage[english]{babel}
\usepackage{enumitem}
\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry}
\usepackage[bottom]{footmisc}
\usepackage[hidelinks]{hyperref}
\usepackage{amsmath}
\usepackage{fixme}
\usepackage{xcolor}

\usepackage{tikz}
\usetikzlibrary{shapes,arrows}

\tikzstyle{block} = [rectangle, draw, fill=orange!20, text width=5em, text centered, minimum height=3em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm, minimum height=3em]

% FIXME Setup
\fxsetup{
  status=draft,
  author=,
  layout=inline,
  theme=color
}
\definecolor{fxnote}{rgb}{0.8000,0.0000,0.0000}
\colorlet{fxnotebg}{yellow}
\makeatletter
\renewcommand*\FXLayoutInline[3]{%
  \@fxdocolon {#3}{%
    \@fxuseface {inline}%
    \colorbox{fx#1bg}{\color {fx#1}\ignorespaces #3\@fxcolon #2}}}
\makeatother

\renewcommand\thesection{\Roman{section}}
\renewcommand\thesubsection{\alph{subsection}}
\titleformat{\section}[block]{\large\scshape\centering}{\thesection.}{1em}{}
\titleformat{\subsection}[block]{\large}{\thesubsection.}{1em}{}

\setlist[itemize]{noitemsep}
\setlist[enumerate]{noitemsep}

\newcommand{\ts}[1]{\mintinline{typescript}{#1}}
\newcommand{\hs}[1]{\mintinline{haskell}{#1}}

\newcommand{\fcy}[1]{\mathcal{#1}}
\newcommand{\lit}[1]{\text{``#1''}}
\newcommand{\etag}[1]{\textsf{#1}}

% TODO: acmart, acmsmall

\title{Typed Interactions for NLU in Opal}
\author{Harrison Goldstein}
\date{Fall 2017}

\begin{document}
\maketitle
\begin{abstract}
  This is an abstract containing relatively abstract things.
\end{abstract}

\section{Introduction} \label{introduction}
In the 1970's, a text-based command interface was ``good enough'' for people
interacting with computers. After all, most of the computation being done was
fairly specific to some application domain, the person operating the computer
was well-trained, and the machine was viewed as a tool for getting the job done.

Today, computers are our personal assistants. They are integrated with our lives
and they help everyday people do everyday tasks. Most people who use computers
today never touch a command shell, and even those who do would often rather not.
Graphical interfaces are amazingly powerful, and are often good enough to allow
users to easily interact with the computer, but even those have their drawbacks.
We would really like our personal assistant to respond to commands in human
language:
\begin{itemize}
\item ``Hey Siri! Play some music.''
\item ``Ok Google. Navigate to my hotel.''
\item ``Slackbot, set my status to active.''
\end{itemize}
This is the realm of Natural Language Understanding (NLU)\footnote{We do not
  distinguish between spoken or written language when talking about NLU in
  general.}.

We have seen a huge growth in NLU applications in recent years, and we will
likely see more. Unfortunately, there has been no serious effort made to make
this power available at the language level, so programmers have been forced to
integrate these technologies into their applications by hand. (In fact, machine
learning in general has been relatively neglected by the programming languages
community.) We propose an addition to the Opal language that will begin to close
this gap.

% \subsection{Opal}
% Opal is a research programming language being developed at Cornell University
% under the guidance of Professor Adrian Sampson. The language is being developed
% as an embedded language within Typescript, and it is designed with machine
% learning applications in mind. Opal supports programming paradigms that fit well
% with the interactive and unpredictable nature of ML programs, and has libraries
% for interacting with various types of ML engines.

\subsection{An Example}
Consider a simple reminder interface to a calendar or other scheduling
assistant. A user might ask the application to set a reminder containing a
message and a time (for simplicity, we will just allow the user to specify
\ts{later} or \ts{tomorrow}) or she
might ask the application to get any active reminders. Since Opal is embedded
in Typescript, a typical Opal application might express this interface with the
following declarations:\footnote{An overview of Typescript's type system can be
  found at \url{www.typescriptlang.org}.}
\begin{minted}{typescript}
  type Msg = string;
  type Time = "later" | "tomorrow";
  type Set = {
    msg: Msg,
    time: Time
  };
  type Intent =
    | { tag: "Set", data: Set }
    | { tag: "Get", data: {} };
\end{minted}

Since we would like to interact with this application via human language, we
need some way of converting a user \emph{utterance} to something of type
\ts{Intent}. Figure \ref{fig:drycleaning} shows the desired
data representation of a user utterance.

\begin{figure}
\begin{minted}{typescript}
  {
    tag: "Set",
    data: {
      msg: "pick up the dry cleaning",
      time: "later",
    }
  }
\end{minted}
  \caption{The desired data representation of the utterance: \emph{Remind me
      later to pick up the dry cleaning}.}
  \label{fig:drycleaning}
\end{figure}

\subsection{Types for NLU}
If one were to try to implement our reminder interface na\"ively, a problem
would arise: the popular backends provide untyped interfaces and require
external configuration. Since Opal hopes to provide a consistent and safe
interface to multiple NLU engines, we decided to develop a typed abstraction for
natural language systems.

In this work, we present a domain specific language (DSL) for simultaneously
specifying
\begin{itemize}
\item configuration for an NLU engine;
\item a typed interface for responses
\end{itemize}
We also implement translations from our DSL to various artifacts which can be
integrated with Opal applications.
\\

We will begin by presenting a technical description of our type language in
section \ref{description}. In section \ref{formalism}, we present a formal
specification for the language our type-directed response parsing algorithm.
Section \ref{implementation} covers the implementation of the language and
related libraries. Finally, sections \ref{evaluation} and \ref{future} present
an evaluation of our work, and our plans for future work, respectively.

\section{Language Description} \label{description}
For our initial work, we elected to focus on building a system to support a
single NLU backend---specifically, we are working with an NLU engine called
Wit.\fxnote{[CITE]} When configuring Wit, as with many NLU backends, the main
task is setting up the \emph{entities} that Wit should be looking for. An entity
is a discrete piece of information that the NLU engine extracts from an
utterance. In our example, the messages and the time are both entites, as is the
user's intent.

\subsection{Search Strategies}
When entities are declared, the user must specify a ``search strategy''. In the
case of Wit, there are three main strategies:
\begin{description}
\item[Free Text] Entities that are found via free text matching are just strings
of text from the utterance. In our example, the message is a free text entity.
\item[Keywords] Keywords entities are just words from some pre-defined set.
  Things like \ts{later} and
  \ts{tomorrow} would be captured via keyword searching.
\item[Trait] These are slightly more abstract. Traits capture something that is
  a quality of the utterance as a whole. Something like a user's intent or even
  mood could be a trait.
\end{description}

To some, these lookup strategies might seem like an implementation detail, but
there is actually something important going on here. The search strategies are
closely tied to the \emph{type} of the desired output. In the case of free text
and keywords, this association is obvious: free text entities are just
\ts{string}s, and keywords are members of a union of literal
types.\footnote{For example, \texttt{"later" | "tomorrow"}.} Traits are slightly
more complex, but they essentially represent a tagged union, where the extracted
entity itself carries only the tag. We will make these ideas more concrete in
section \ref{formalism}.

\subsection{The DSL}
With our observations about search strategies in mind, we can write our types
from before in a slightly different way. Here are the same type declarations
from section \ref{introduction}, but written in our DSL:
\begin{minted}{typescript}
  free-text Msg;
  keywords Time = "later" | "tomorrow";
  alias Set = {
    msg: Msg,
    time: Time
  };
  trait Intent =
    | <Set> Set
    | <Get> {};
\end{minted}

The main thing to notice is that the keyword \ts{type} has been replaced by
entity declarations \ts{free-text}, \ts{keywords}, and \ts{trait}, as well as
\ts{alias}. There are a number of benefits of entity declarations. First and
foremost, it allows the user to explicitly define the NLU engine configuration
alongside the program types. With types defined in our DSL, we can automatically
generate configuration for entities that is guaranteed to be consistent with the
types in the program. When a \ts{Time} is returned to the user, it will be
guaranteed to be either \ts{"later"} or \ts{"tomorrow"}. This alone solves a lot
of the problems that tend to come up when writing these types of applications.

Another benefit of these bespoke declaration forms is simplified syntax. When a
\ts{free-text} entity is declared, there is no need to write ``\ts{= string}'',
because that is implied by the search strategy. Similarly, since traits are
always tagged unions, we can write them in (almost) ML syntax, rather than
writing out the idiomatic Typescript workaround.

Finally, writing specific entity declarations rather than just \ts{type}, makes
it clear that there is more going on than a direct translation into typescript.
In fact, the DSL parser does lot of validation to guarantee that only reasonable
types can be written down. For example, a \ts{keywords} declaration is required
to be a union, and that union must contain only literals. Even the \ts{alias}
keyword has some validation built in. Some Typescript types would be problematic
if they were allowed---we enforce those conditions in the DSL.

\begin{figure}
\begin{minted}{haskell}
  data Decl = FreeText String
            | Keywords String [String]
            | Trait String [(String, Ty)]
            | Alias String Ty
            deriving (Show)

  data Ty = Def String
          | Lit String
          | Rec [(String, Ty)]
          deriving (Show)
\end{minted}
  \caption{The abstract syntax tree for our DSL.}
  \label{fig:ast}
\end{figure}

Figure \ref{fig:ast} shows the AST that is produced when we parse the DSL.
Looking at this, it also becomes clear that the limitations on the types allowed
in a trait declaration are fairly strict. As of writing, there are only three
types allowed within a trait declaration: previously defined entities, literals,
and records (where each element is subject to the same restrictions). These
limitations match up with the capabilities of NLU engines in general. While many
backends provide the ability to parse numbers, dates, and other data, these
types of entities are implementation specific. In section \ref{future}, we
discuss plans to expand the allowed types in a modular way.

\subsection{NLU Engine Responses}
Our DSL solves quite a few problems on its own. It forces the program types to
agree with the NLU configuration, and it provides a convenient way to write both
down together. It also prevents the user from having unreasonable expectations
of the backend, by restricting the types that a user can expect as a response.

Unfortunately, there is still one major issue. The vast majority of NLU engines
don't return structured data. Instead, they return a bag of entities with no
information about the relationships between them. Figure \ref{fig:nostructure}
shows a sample response. Recall that in our example, \ts{Msg} and \ts{Time}
entities were only expected when there was an \ts{Intent} with tag \ts{"Set"}.
Furthermore, our types dictate exactly where in the data structure those entity
values should end up. The popular libraries, Wit incuded, igore this structure
entirely.

\begin{figure}
\begin{minted}{typescript}
  {
    Intent: "Set",
    Msg: "pick up the dry cleaning",
    Time: "later"
  }
\end{minted}
  \caption{The unstructured response for the utterance: \emph{Remind me later to
      pick up the dry cleaning}.}
  \label{fig:nostructure}
\end{figure}

In order to reconstruct the desired data from the NLU backend's response, we
realized that it was necessary to keep data from the DSL around at run-time, not
just during setup.

\begin{figure*}
  \centering
  \begin{tikzpicture}[node distance = 3cm, auto]
    \node [block] (types) {Type DSL};
    \node [block, right of=types] (decls_hs) {Decl (hs)};
    \node [block, right of=decls_hs] (decls_ts) {Runtypes (ts)};
    \node [block, below of=decls_ts] (cfg) {NLU Config};
    \node [block, above of=decls_ts] (ts_types) {Types (ts)};
    \node [cloud, right of=decls_ts] (ts) {Opal};
    \node [cloud, right of=cfg] (backend) {Wit};

    \path [line] (types) -- (decls_hs);
    \path [line] (decls_hs) -- (decls_ts);
    \path [line] (decls_hs) -- (cfg);
    \path [line] (decls_hs) -- (ts_types);
    \path [line] (decls_ts) -- (ts);
    \path [line] (ts_types) -- (ts);
    \path [line] (cfg) -- (backend);
  \end{tikzpicture}
  \caption{The configuration process.}
  \label{fig:process}
\end{figure*}

We elected to use a typescript library called \emph{runtypes} to serialize the
AST directly into typescript; this gives us a data structure to guide response
parsing, and also provides us a way to dynamically type-check the results. We
discuss the response parsing algorithm in more detail in section
\ref{implementation}.

Figure \ref{fig:process} shows process of configuring an application for NLU
interaction. The programmer writes types in our DSL, and uses our Haskell tool
to parse the declarations. The tool serializes three artifacts: a set of
Typescript type declarations, a runtypes datastructure, and a set of JSON
configuration files for Wit. The programmer simply imports the types and
runtypes into their Opal project, uploads the JSON files to Wit, and the system
is ready to go.

\section{Formalism} \label{formalism}
In order to reason about the correctness of our translations, we have formalized
the results in section \ref{description}. We begin by formalizing the type DSL
as a pseuo-BNF grammar. This is shown in figure \ref{fig:grammar}.

\begin{figure}
  \centering
  \begin{align*}
    \fcy{D} ::=&\ \text{FreeText}(\etag{t}) \\
    |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
    |&\ \text{Trait}(\etag{t},\ \{s: \tau\}) \\
    |&\ \text{Alias}(\etag{t},\ \tau) \\
    \\
    \tau ::=&\ \etag{t} \mid \ell \mid \{s: \tau\} \mid \dots
  \end{align*}
  $$ \etag{t} \in \textsf{Tag} \qquad \ell \in \textsf{Literal} \qquad s \in
  \textbf{string} $$
  \caption{A grammar for the DSL.}
  \label{fig:grammar}
\end{figure}

The translation from the haskell declarations above is mostly straightforward,
but there are a few notable differences. First, we have separated out tags,
literals, and strings. This is mostly for notational convenience, but these
things are also slightly different when serialized to Typescript. Tags become
names in Typescript, identifying a specific type alias. Literals become strings
in a literal type. Finally, strings become simple strings, used within a record
as keys.

We will define a set $\textsf{Entity} \subseteq \textsf{Tag}$ to represent the
available entity tags. Any of these might be needed to reconstruct the
appropriate type from the response.

Given a tag \etag{t}, we can define a parser in terms of \etag{t} as follows:
\begin{align*}
  \text{interp}\ &:\ \textsf{Tag} \rightarrow \fcy{T} \\
  \text{parse}\ &:\ 2^{\textsf{Entity}} \rightharpoonup \text{interp}(\etag{t})
\end{align*}
The interp map computes the Typescript type of \etag{t}. Then, we can represent
parsing as a partial function from sets of entities to typescript objects of the
correct type.

\section{Implementation} \label{implementation}
Explain the architecture of the implementation. Discuss how the tools should be
used.

\section{Evaluation} \label{evaluation}
Talk about how great we did.

\section{Future Work} \label{future}
{\bf Brain Dump}
\begin{itemize}
\item More types (include Wit builtins)
\item More backends
\item Use confidence information
\item Unit tests as training samples
\end{itemize}

\section{Conclusion} \label{conclusion}

\end{document}

### 2017-12-08 test.tex
Initial commit

\documentclass[format=acmsmall, review=false, screen=true]{acmart}

\usepackage{booktabs} % For formal tables

\usepackage[ruled]{algorithm2e} % For algorithms
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
\acmJournal{TWEB}
\acmVolume{9}
\acmNumber{4}
\acmArticle{39}
\acmYear{2010}
\acmMonth{3}
\copyrightyear{2009}
%\acmArticleSeq{9}

% Copyright
%\setcopyright{acmcopyright}
\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

% DOI
\acmDOI{0000001.0000001}

% Paper history
\received{February 2007}
\received[revised]{March 2009}
\received[accepted]{June 2009}

% Document starts
\begin{document}
% Title portion. Note the short title for running heads
\title[A Multifrequency MAC for Wireless Sensor]{A Multifrequency MAC
  Specially Designed for Wireless Sensor  Network Applications}

\author{Gang Zhou}
\orcid{1234-5678-9012-3456}
\affiliation{%
  \institution{College of William and Mary}
  \streetaddress{104 Jamestown Rd}
  \city{Williamsburg}
  \state{VA}
  \postcode{23185}
  \country{USA}}
\email{gang_zhou@wm.edu}
\author{Valerie B\'eranger}
\affiliation{%
  \institution{Inria Paris-Rocquencourt}
  \city{Rocquencourt}
  \country{France}
}
\email{beranger@inria.fr}
\author{Aparna Patel}
\affiliation{%
 \institution{Rajiv Gandhi University}
 \streetaddress{Rono-Hills}
 \city{Doimukh}
 \state{Arunachal Pradesh}
 \country{India}}
\email{aprna_patel@rguhs.ac.in}
\author{Huifen Chan}
\affiliation{%
  \institution{Tsinghua University}
  \streetaddress{30 Shuangqing Rd}
  \city{Haidian Qu}
  \state{Beijing Shi}
  \country{China}
}
\email{chan0345@tsinghua.edu.cn}
\author{Ting Yan}
\affiliation{%
  \institution{Eaton Innovation Center}
  \city{Prague}
  \country{Czech Republic}}
\email{yanting02@gmail.com}
\author{Tian He}
\affiliation{%
  \institution{University of Virginia}
  \department{School of Engineering}
  \city{Charlottesville}
  \state{VA}
  \postcode{22903}
  \country{USA}
}
\affiliation{%
  \institution{University of Minnesota}
  \country{USA}}
\email{tinghe@uva.edu}
\author{Chengdu Huang}
\author{John A. Stankovic}
\author{Tarek F. Abdelzaher}
\affiliation{%
  \institution{University of Virginia}
  \department{School of Engineering}
  \city{Charlottesville}
  \state{VA}
  \postcode{22903}
  \country{USA}
}

\begin{abstract}
Multifrequency media access control has been well understood in
general wireless ad hoc networks, while in wireless sensor networks,
researchers still focus on single frequency solutions. In wireless
sensor networks, each device is typically equipped with a single
radio transceiver and applications adopt much smaller packet sizes
compared to those in general wireless ad hoc networks. Hence, the
multifrequency MAC protocols proposed for general wireless ad hoc
networks are not suitable for wireless sensor network applications,
which we further demonstrate through our simulation experiments. In
this article, we propose MMSN, which takes advantage of
multifrequency availability while, at the same time, takes into
consideration the restrictions of wireless sensor networks. Through
extensive experiments, MMSN exhibits the prominent ability to utilize
parallel transmissions among neighboring nodes.
\end{abstract}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%
% End generated code
%

\keywords{Wireless sensor networks, media access control,
multi-channel, radio interference, time synchronization}

\maketitle

% The default list of authors is too long for headers.
\renewcommand{\shortauthors}{G. Zhou et al.}

\end{document}

### 2017-12-11 paper.tex
Added content, yay

@@
 \usepackage{xcolor}
+\usepackage{subcaption}
 
@@
 In order to reason about the correctness of our translations, we have formalized
-the results in section \ref{description}. We begin by formalizing the type DSL
-as a pseuo-BNF grammar. This is shown in figure \ref{fig:grammar}.
-
+the results from section \ref{description}. We begin by formalizing the type DSL
+as a pseuo-BNF grammar. This is shown in figure \ref{fig:grammar_a}.
 \begin{figure}
   \centering
+  \begin{subfigure}{1\linewidth}
     \begin{align*}
@@
       |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
-    |&\ \text{Trait}(\etag{t},\ \{s: \tau\}) \\
-    |&\ \text{Alias}(\etag{t},\ \tau) \\
+      |&\ \text{Trait}(\etag{t},\ \{s: \fcy{T}\}) \\
       \\
-    \tau ::=&\ \etag{t} \mid \ell \mid \{s: \tau\} \mid \dots
+      \fcy{T} ::=&\ \fcy{D} \mid \ell \mid \{s: \fcy{T}\}
     \end{align*}
-  $$ \etag{t} \in \textsf{Tag} \qquad \ell \in \textsf{Literal} \qquad s \in
+    $$ \etag{t} \in \textsf{Tag} \qquad \ell, s \in \textbf{string} $$
+    \caption{Type DSL}
+    \label{fig:grammar_a}
+  \end{subfigure}\vspace{1cm}
+  \begin{subfigure}{1\linewidth}
+    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \ell
+    \rangle $$
+    \caption{Relevant Opal Types}
+    \label{fig:grammar_b}
+  \end{subfigure}
+  \caption{Formal grammars.}
+\end{figure}
+There are similarities between these definitions and the haskell ones in figure
+\ref{fig:ast}, but it has some slightly different goals. First, we remove
+aliases and require that all type definitions are written ``in-line''. Next,
+rather than split the definition by allowing tags within $\fcy{T}$, we also
+require that entitiy definitions are inlined. This means that the data a user
+expects from the NLU backend can be represented by a single tree of type
+$\fcy{D}$. Importantly, none of these simplifications are fundamental---they
+simply remove some aspects of the system that make it easier to use in practice.
+
+In addition to a grammar for the DSL itself, in figure \ref{fig:grammar_b} we
+define a grammar of types, $\tau$ which will represent Opal types specifically.
+We can think of $\tau$ as representing the final form of the data that a user
+hopes to obtain, with no information about the entities that generated the data.
+We add a few types that are noteworthy: the type {\bf string} is added to
+account for FreeText entities; $\langle {\ell} \rangle$ represents a disjunction
+of literals, and covers the case of Keywords entities.
+
+As mentioned above, user's view of their system can be formalized as a top-level
+declaration generated by $\fcy{D}$. Since this declaration encompases both type
+and configuration information, we would like a way to compute just the
+\emph{type} of data that the user expects from the NLU engine. We define the
+following mutually recursive functions, which traverse a declaration and
+construct a single type in $\tau$.
+\begin{align*}
+  \text{interp}\ &:\ \fcy{D} \to \tau \\
+  \text{interpTy}\ &:\ \fcy{T} \to \tau
+\end{align*}
+
+With the basic building blocks defined, we can move forward in formalizing our
+system. \fxnote{Should I write out implementations?}
+
+\subsection{Configuration Generation}
+The first step that our system takes is configuration. In practice, this is a
+two-part process: the user needs to extract the appropriate Opal code as well as
+the configuration for the NLU backend. The first step is covered by our
+definition of interp. Since we are taking $\tau$ as an analog for relevant Opal
+types, the type generation process is exactly the execution of interp on the
+top-level entity declaration.
+
+As for the second step, we would like the NLU backend to be set up such that its
+responses are \emph{well-formed} with respect to the configuration. Intuitively,
+any time we look up some entity tag in the response, we would like to get data
+that is compatible with that entity's declaration. We define a type,
+\textsf{Resp} of responses, and require that a lookup function exists, with the
+following type:
+$$ \text{lookup}\ :\ \textsf{Tag} \to \textsf{Resp} \rightharpoonup
 \textbf{string} $$
-  \caption{A grammar for the DSL.}
-  \label{fig:grammar}
-\end{figure}
+Note that the function is partial; there is no guarantee that the response will
+contain the desired data. In the sequel, we will often assume that lookup is
+defined on tags we care about, but that assumption will be stated explicitly
+rather than assumed by default.
 
-The translation from the haskell declarations above is mostly straightforward,
-but there are a few notable differences. First, we have separated out tags,
-literals, and strings. This is mostly for notational convenience, but these
-things are also slightly different when serialized to Typescript. Tags become
-names in Typescript, identifying a specific type alias. Literals become strings
-in a literal type. Finally, strings become simple strings, used within a record
-as keys.
+The lookup function always returns a string, but the string might mean any
+number of things. The data might be a particular literal, in the case of a
+keywords or trait entity, or an entire phrase in the case of a free-text entity.
+We define configurations as functions that map a subset of entity tags to the
+type of data expected for those entities.
+$$ \textsf{Cfg} \equiv \textsf{Tag} \rightharpoonup \tau $$
+Now, we call a response well-formed with respect to $\text{cfg} : \textsf{Cfg}$
+if and only if it returns appropriate data on lookup. More formally, for some $r
+: \textsf{Resp}$,
+\begin{align*}
+  &r\ \text{well-formed}_{\text{cfg}} \iff \\
+  &\forall (\etag{t} : \textsf{Tag}).\ \text{cfg}(\etag{t})\ \text{defined} \Rightarrow \text{lookup}(\etag{t}, r) : \text{cfg(\etag{t})}
+\end{align*}
 
-We will define a set $\textsf{Entity} \subseteq \textsf{Tag}$ to represent the
-available entity tags. Any of these might be needed to reconstruct the
-appropriate type from the response.
+\subsection{Response Parsing}
+We discussed above that there is a challenge in parsing the unstructured
+response from the NLU engine into data of the appropriate types. This section
+will formalize that problem and present our solution.
 
-Given a tag \etag{t}, we can define a parser in terms of \etag{t} as follows:
+Our goal will be to write a parsing function which takes a declaration, $d$, and
+a response, and produces a value of type interp($d$). We will again write two
+mutually recursive functions, parse and parseTy, with types:
 \begin{align*}
-  \text{interp}\ &:\ \textsf{Tag} \rightarrow \fcy{T} \\
-  \text{parse}\ &:\ 2^{\textsf{Entity}} \rightharpoonup \text{interp}(\etag{t})
+\text{parse}\ &:\ \Pi_{(d : \fcy{D})}. \textsf{Resp} \rightharpoonup \text{interp}(d) \\
+\text{parseTy}\ &:\ \Pi_{(t : \fcy{T})}. \textsf{Resp} \rightharpoonup \text{interp}(t)
 \end{align*}
-The interp map computes the Typescript type of \etag{t}. Then, we can represent
-parsing as a partial function from sets of entities to typescript objects of the
-correct type.
+Since we need the type of the result to depend on the declaration, these are
+dependent functions. The functions are also partial because lookups might fail.
+
+We will define parse by case analysis on the declaration.
+\begin{align*}
+  \text{parse}(\text{FreeText}(\etag{t}), r) = \text{lookup}(\etag{t}, r)
+\end{align*}
+This case is fairly straightforward. Since the interpretation of a free-text
+entity is just a string, we can simply look up the value of the appropriate
+entity from the response.
+\begin{align*}
+  \text{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
+  \text{lookup}(\etag{t}, r)
+\end{align*}
+Perhaps surprisingly, we can use the exact same definition in the Keywords case
+that we did in the FreeText case. We are required to return a value of type
+$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$, but since the NLU engine has
+been configured with the same list of literals, we are guaranteed to get one of
+them back from lookup (assuming lookup is defined).
+\begin{align*}
+  \text{parse}(\text{Trait}(\etag{t}, rec), r) = \\
+  \text{parseTy}(rec[\text{lookup}(\etag{t}, r)], r)
+\end{align*}
+The final case for parse simply matches the value in the response to an index
+into the record, and interprets that record entry using parseTy.
+
+We will not write out the full definition of parseTy, but note that it is fairly
+automatic. If the type is another declaration, we can call back to parse, if it
+is a literal we simply generate the appropriate string, and if it is a record,
+we build the appropriate record and recursively build out its fields.
+
+\subsection{Correctness}
+Given a declaration $d$, we generate a configuration, cfg. Then,
+\begin{flalign*}
+  &\forall\ (r : \textsf{Resp}).\\
+  &r\ \text{well-formed}_{\text{cfg}} \Rightarrow \text{parse}(d, r)\ \text{defined}
+\end{flalign*}
 
 \section{Implementation} \label{implementation}
-Explain the architecture of the implementation. Discuss how the tools should be
-used.
+
+\subsection{Configuration Tool}
+When a user wants to configure an Opal application to interact with an NLU
+library like Wit, they begin by using our configuration tool. The tool turns
+code in the type DSL into typescript code and JSON configuration, and ensures
+that the NLU engine responses will agree with the type definitions in the user's
+application.
+
+Our configuration tool is written in Haskell, and is designed to be simple and
+obviously correct. We use the Parsec library to parse the type DSL into the
+intermediate representation mentioned above in figure \ref{fig:ast}.
+
+\subsection{Runtime Library}
 

### 2017-12-12 paper.tex
Added questions document, more additions to the paper

@@
 
-% TODO: acmart, acmsmall
-
 \title{Typed Interactions for NLU in Opal}
@@
 \maketitle
+
 \begin{abstract}
-  This is an abstract containing relatively abstract things.
+  Natural Language Understanding (NLU) is a powerful tool for modern
+  applications. Interacting with remote NLU engines can be error prone: most
+  backends provide no type guarantees, and configuration synchronization is a
+  constant source of bugs. We present a DSL for configuring a NLU application
+  that ensures synchronization and type-safety.
 \end{abstract}
@@
 this gap.
-
-% \subsection{Opal}
-% Opal is a research programming language being developed at Cornell University
-% under the guidance of Professor Adrian Sampson. The language is being developed
-% as an embedded language within Typescript, and it is designed with machine
-% learning applications in mind. Opal supports programming paradigms that fit well
-% with the interactive and unpredictable nature of ML programs, and has libraries
-% for interacting with various types of ML engines.
 
@@
 With our observations about search strategies in mind, we can write our types
-from before in a slightly different way. Here are the same type declarations
-from section \ref{introduction}, but written in our DSL:
+from before in a slightly different way. Figure \ref{fig:dsl_decls} same type
+declarations from section \ref{introduction} written in our DSL.
+
+\begin{figure}
 \begin{minted}{typescript}
@@
 \end{minted}
+  \caption{Some DSL declarations.}
+  \label{fig:dsl_decls}
+\end{figure}
 
@@
 Unfortunately, there is still one major issue. The vast majority of NLU engines
-don't return structured data. Instead, they return a bag of entities with no
+do not return structured data. Instead, they return a bag of entities with no
 information about the relationships between them. Figure \ref{fig:nostructure}
@@
 that we did in the FreeText case. We are required to return a value of type
-$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$, but since the NLU engine has
-been configured with the same list of literals, we are guaranteed to get one of
-them back from lookup (assuming lookup is defined).
+$\langle {\ell_1 \mid \dots \mid \ell_n} \rangle$; as long as $r$ is
+well-formed, it will have been configured to have the same list of literals. We
+are guaranteed to get one of them back from lookup.
 \begin{align*}
@@
 \end{flalign*}
+Note that since parse is a dependent function, it will already return a value of
+the correct type. Thus, we only need to care that it is defined. Proof of this
+fact is clear by induction on $d$.
 
@@
 \subsection{Runtime Library}
+At run-time typical application using NLU will go through a fairly consistent
+cycle.
+\begin{enumerate}
+\item Ask user for input, converting voice to text if necessary.
+\item Query the NLU engine and obtain a response for the given input.
+\item Parse the response into application data.
+\item Process parsed data, and repeat.
+\end{enumerate}
+Steps 1 and 4 are generally application-specific, so we have elected to provide
+a library which handles steps 2 and 4. For each backend that Opal
+supports,\footnote{At the time of writing, Wit is the only supported backend.}
+we provide utilities for querying the API and parsing the response data. The
+network code involved in obtaining the response is fairly uninteresting, so we
+will focus on the parsing,
+
+In section \ref{formalism}, we outline the parse function, which takes a
+declaration and a response and generates an Opal object. The function that we
+provide in the Opal library is similar, but there are two important differences.
+First, the parse function takes some element of $\fcy{D}$, which it uses to
+deduce the structure of the desired output. In the real application, the analog
+of $\fcy{D}$ is the \hs{Decl} type in the Haskell code; this is not available to
+the Opal application at run time. As mentioned above, we get around this by
+using a modified version the runtypes library to keep this information around at
+run-time. We extend runtypes to include a type for entities---which map to
+declarations in $\fcy{D}$. Runtypes provides the infrastructure for non-entity
+types in $\fcy{T}$.
+
+The other difference arises because Opal does not have dependent types. We
+cannot completely faithfully transcribe the parse function from above, because
+it is not obvious that it can be typed at all. The solution here is to combine
+Opal's dynamic typing with runtypes' dynamic checking feature to create a
+function that appears to be dependently typed. First, we use dynamic types
+within the parse function. While building the response data, we assume, where
+necessary, that it has type \ts{any}. Then, once parsing is complete, we use the
+runtypes object to check the parsed object, and we cast it to the desired type.
+The definition is shown in figure \ref{fig:parse}. The cast is safe, so long as
+the user respects the invariant that parse is always called with compatible type
+and runtype parameters.
+
+\begin{figure*}
+\begin{minted}{typescript}
+          public parse<T>(rt: Runtype<any>, res: Response): T {
+              return rt.check(this._parse(rt, res)) as T;
+          }
+
+          private _parse(rt: Runtype<any>, res: Response): any { ... }
+\end{minted}
+  \caption{The wit parse functions.}
+  \label{fig:parse}
+\end{figure*}
 
 \section{Evaluation} \label{evaluation}
-Talk about how great we did.
+%TODO: Cover limitations of the design, possible use-cases, etc
 
 \section{Future Work} \label{future}
-{\bf Brain Dump}
-\begin{itemize}
-\item More types (include Wit builtins)
-\item More backends
-\item Use confidence information
-\item Unit tests as training samples
-\end{itemize}
+Moving forward, we hope to take this work in two main directions---wider support
+for NLU features and backends and entirely new features.
+
+\subsection{Wider Support}
+Expanding support for more entity types and backends is an obvious way to
+increase the usefulness of our work. Here, we discuss both of those extensions
+in detail.
+
+\subsubsection{More Types}
+In addition to the three ``universal'' kinds of entities (free-text, keywords,
+and traits), the Wit NLU engine provides more specialized ways of looking up
+entities. Wit provides support for parsing numbers, dates, times, and other
+specific forms of text. We elected to keep things simple at first and exclude
+these built-in entities, but adding support for them seems like a very
+reasonable next-step.
+
+At a first approximation, this would require modifying the definitions of
+$\fcy{D}$ and $\tau$ to include declarations for built-in entities and
+base types for the data associated with those entities. More work needs to be
+done exploring this space before any more specific conclusions can be drawn.
+
+\subsubsection{More Backends}
+We are also interested exploring NLU engines beyond Wit. One reason for this is
+purely practical---users might want to use other backends, and it would be nice
+for Opal to support this. Another reason, however, is more academic. Exploring
+other NLU engines would give us opportunities to test the generality of our
+current design. Preliminary research suggests that building our type system
+around entities and search strategies makes sense, but we can only know that for
+sure if we try to build interfaces for more engines.
+
+\subsection{Interesting Features}
+There are a number of novel research directions available as extensions to this
+project. The following are two of the many ideas that we have had.
+
+\subsubsection{Confidence-Based Parsing}
+Up to this point, we have assumed very little about the data returned in the NLU
+response. In practice, however, many NLU engines provide more than just
+information about the matched entities. Wit, in particular, provides confidence
+levels for each entitiy that it finds. Thus, a value like \ts{"Set"} might be
+paired with a confidence level of $70\%$, meaning that Wit is $70\%$ sure that
+the user's intent is actually to set a new reminder.
+
+As of now, we do not actually use this information. In the majority of cases
+that we have tested, a Wit bot either has enough information to get the right
+entities, or it does not. However, it is possible that as we attempt to support
+new types of entities and as user types get more complicated, it will be useful
+to try multiple parses of a single response. In that case, we can choose between
+different parses using confidence scores.
+
+\subsubsection{Training with Tests}
+A final extension of this project would attempt to close the code/configuration
+gap even further by allowing users to train their NLU application in a way that
+is guaranteed to support their tests. We would accomplish this using the same
+idea behind our type DSL---we introduce a new DSL (or an extension to the
+current one) for generating both Wit training examples and application tests.
+While synchronization is less of an issue with training than it is with
+configuration, it would still be extremely convenient to express the entirety of
+the application's behavior in one place.
+
+Exploring this path would require a fair amount of implementation work, but we
+would also need to take great care to avoid overfitting. It is unclear to what
+extent this would become an issue.
 
 \section{Conclusion} \label{conclusion}
+% TODO: Write
 

### 2017-12-13 paper.tex
Changes to formalism section
- Added nicer text for function definitions
- Revisions of wording, etc

@@
 \newcommand{\etag}[1]{\textsf{#1}}
+
+\newcommand{\ff}[1]{\textsf{#1}}
 
@@
 Finally, writing specific entity declarations rather than just \ts{type}, makes
-it clear that there is more going on than a direct translation into typescript.
+it clear that there is more going on than a direct translation into Typescript.
 In fact, the DSL parser does lot of validation to guarantee that only reasonable
@@
 
-We elected to use a typescript library called \emph{runtypes} to serialize the
-AST directly into typescript; this gives us a data structure to guide response
+We elected to use a Typescript library called \emph{runtypes} to serialize the
+AST directly into Typescript; this gives us a data structure to guide response
 parsing, and also provides us a way to dynamically type-check the results. We
@@
 the results from section \ref{description}. We begin by formalizing the type DSL
-as a pseuo-BNF grammar. This is shown in figure \ref{fig:grammar_a}.
+as shown in figure \ref{fig:grammar_a}.
 \begin{figure}
@@
     \end{align*}
-    $$ \etag{t} \in \textsf{Tag} \qquad \ell, s \in \textbf{string} $$
+    $$ \etag{t} \in \ff{Tag} \qquad \ell, s \in \textbf{string} $$
     \caption{Type DSL}
@@
 \end{figure}
-There are similarities between these definitions and the haskell ones in figure
-\ref{fig:ast}, but it has some slightly different goals. First, we remove
+There are similarities between these definitions and the Haskell ones in figure
+\ref{fig:ast}, but these have some slightly different goals. First, we remove
 aliases and require that all type definitions are written ``in-line''. Next,
@@
 expects from the NLU backend can be represented by a single tree of type
-$\fcy{D}$. Importantly, none of these simplifications are fundamental---they
-simply remove some aspects of the system that make it easier to use in practice.
+$\fcy{D}$. Importantly, none of these simplifications are fundamental---this
+formulation can represent the same types as the actual system. This approach
+simply gives up some usability to simplify reasoning.
 
@@
 \begin{align*}
-  \text{interp}\ &:\ \fcy{D} \to \tau \\
-  \text{interpTy}\ &:\ \fcy{T} \to \tau
+  \ff{interp}\ &:\ \fcy{D} \to \tau \\
+  \ff{interpTy}\ &:\ \fcy{T} \to \tau
 \end{align*}
@@
 the configuration for the NLU backend. The first step is covered by our
-definition of interp. Since we are taking $\tau$ as an analog for relevant Opal
-types, the type generation process is exactly the execution of interp on the
-top-level entity declaration.
+definition of \ff{interp}. Since we are taking $\tau$ as an analog for relevant
+Opal types, the type generation process is exactly the execution of \ff{interp}
+on the top-level entity declaration.
 
@@
 that is compatible with that entity's declaration. We define a type,
-\textsf{Resp} of responses, and require that a lookup function exists, with the
-following type:
-$$ \text{lookup}\ :\ \textsf{Tag} \to \textsf{Resp} \rightharpoonup
-\textbf{string} $$
+\textsf{Resp} of responses, and require that a \ff{lookup} function exists, with
+the following type:
+$$ \ff{lookup}\ :\ \ff{Tag} \to \ff{Resp} \rightharpoonup \textbf{string} $$
 Note that the function is partial; there is no guarantee that the response will
-contain the desired data. In the sequel, we will often assume that lookup is
-defined on tags we care about, but that assumption will be stated explicitly
+contain the desired data. In the sequel, we will often assume that \ff{lookup}
+is defined on tags we care about, but that assumption will be stated explicitly
 rather than assumed by default.
 
-The lookup function always returns a string, but the string might mean any
+The \ff{lookup} function always returns a string, but the string might mean any
 number of things. The data might be a particular literal, in the case of a
@@
 type of data expected for those entities.
-$$ \textsf{Cfg} \equiv \textsf{Tag} \rightharpoonup \tau $$
-Now, we call a response well-formed with respect to $\text{cfg} : \textsf{Cfg}$
-if and only if it returns appropriate data on lookup. More formally, for some $r
-: \textsf{Resp}$,
+$$ \ff{Cfg} \equiv \ff{Tag} \rightharpoonup \tau $$
+Now, we call a response well-formed with respect to $\ff{cfg} : \ff{Cfg}$ if and
+only if it returns appropriate data on \ff{lookup}. More formally, for some $r :
+\ff{Resp}$,
 \begin{align*}
   &r\ \text{well-formed}_{\text{cfg}} \iff \\
-  &\forall (\etag{t} : \textsf{Tag}).\ \text{cfg}(\etag{t})\ \text{defined} \Rightarrow \text{lookup}(\etag{t}, r) : \text{cfg(\etag{t})}
+  &\forall (\etag{t} : \textsf{Tag}).\ \ff{cfg}(\etag{t})\ \text{defined} \Rightarrow \ff{lookup}(\etag{t}, r) : \ff{cfg}(\etag{t})
 \end{align*}
@@
 Our goal will be to write a parsing function which takes a declaration, $d$, and
-a response, and produces a value of type interp($d$). We will again write two
-mutually recursive functions, parse and parseTy, with types:
+a response, and produces a value of type \ff{interp}($d$). We will again write
+two mutually recursive functions, \ff{parse} and \ff{parseTy}, with types:
 \begin{align*}
-\text{parse}\ &:\ \Pi_{(d : \fcy{D})}. \textsf{Resp} \rightharpoonup \text{interp}(d) \\
-\text{parseTy}\ &:\ \Pi_{(t : \fcy{T})}. \textsf{Resp} \rightharpoonup \text{interp}(t)
+\ff{parse}\ &:\ \Pi_{(d : \fcy{D})}. \textsf{Resp} \rightharpoonup \ff{interp}(d) \\
+\ff{parseTy}\ &:\ \Pi_{(t : \fcy{T})}. \textsf{Resp} \rightharpoonup \ff{interp}(t)
 \end{align*}
@@
 
-We will define parse by case analysis on the declaration.
+We will define \ff{parse} by case analysis on the declaration.
 \begin{align*}
-  \text{parse}(\text{FreeText}(\etag{t}), r) = \text{lookup}(\etag{t}, r)
+  \ff{parse}(\text{FreeText}(\etag{t}), r) = \ff{lookup}(\etag{t}, r)
 \end{align*}
 This case is fairly straightforward. Since the interpretation of a free-text
-entity is just a string, we can simply look up the value of the appropriate
-entity from the response.
+entity is just a {\bf string}, we can simply look up the value of the
+appropriate entity from the response.
 \begin{align*}
-  \text{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
-  \text{lookup}(\etag{t}, r)
+  \ff{parse}(\text{Keywords}(\etag{t}, [\ell_1, \dots, \ell_n]), r) =\\
+  \ff{lookup}(\etag{t}, r)
 \end{align*}
@@
 well-formed, it will have been configured to have the same list of literals. We
-are guaranteed to get one of them back from lookup.
+are guaranteed to get one of them back from \ff{lookup}.
 \begin{align*}
-  \text{parse}(\text{Trait}(\etag{t}, rec), r) = \\
-  \text{parseTy}(rec[\text{lookup}(\etag{t}, r)], r)
+  \ff{parse}(\text{Trait}(\etag{t}, rec), r) = \\
+  \ff{parseTy}(rec[\ff{lookup}(\etag{t}, r)], r)
 \end{align*}
-The final case for parse simply matches the value in the response to an index
-into the record, and interprets that record entry using parseTy.
+The final case for \ff{parse} simply matches the value in the response to an
+index into the record, and interprets that record entry using \ff{parseTy}.
 
-We will not write out the full definition of parseTy, but note that it is fairly
-automatic. If the type is another declaration, we can call back to parse, if it
-is a literal we simply generate the appropriate string, and if it is a record,
-we build the appropriate record and recursively build out its fields.
+We will not write out the full definition of \ff{parseTy}, but note that it is
+fairly automatic. If the type is another declaration, we can call back to
+\ff{parse}, if it is a literal we simply generate the appropriate string, and if
+it is a record, we build the appropriate record and recursively build out its
+fields.
 
 \subsection{Correctness}
-Given a declaration $d$, we generate a configuration, cfg. Then,
+Given a declaration $d$, we generate a configuration, \ff{cfg}. Then,
 \begin{flalign*}
@@
 \end{flalign*}
-Note that since parse is a dependent function, it will already return a value of
-the correct type. Thus, we only need to care that it is defined. Proof of this
-fact is clear by induction on $d$.
+Note that since \ff{parse} is a dependent function, it will already return a
+value of the correct type. Thus, we only need to care that it is defined. Proof
+of this fact is clear by induction on $d$.
 
@@
 library like Wit, they begin by using our configuration tool. The tool turns
-code in the type DSL into typescript code and JSON configuration, and ensures
+code in the type DSL into Typescript code and JSON configuration, and ensures
 that the NLU engine responses will agree with the type definitions in the user's
@@
 \section{Conclusion} \label{conclusion}
-% TODO: Write
+The Opal langauge hopes to be the ideal language for building modern
+applications that leverage machine learning techniques. As NLU becomes more and
+more popular as a user interface, it makes sense that Opal would provide a clean
+and safe interface for natural langauge interaction.
 

### 2017-12-13 paper.tex
Cleaned up part of the formalism

@@
 \usepackage{amsmath}
+\usepackage{amsthm}
 \usepackage{fixme}
@@
 \newcommand{\ff}[1]{\textsf{#1}}
+
+\newtheorem{theorem}{Theorem}
 
@@
 In addition to a grammar for the DSL itself, in figure \ref{fig:grammar_b} we
-define a grammar of types, $\tau$ which will represent Opal types specifically.
-We can think of $\tau$ as representing the final form of the data that a user
-hopes to obtain, with no information about the entities that generated the data.
-We add a few types that are noteworthy: the type {\bf string} is added to
-account for FreeText entities; $\langle {\ell} \rangle$ represents a disjunction
-of literals, and covers the case of Keywords entities.
+define a grammar of types, $\tau$. We can think of $\tau$ as representing the
+final form of the data that a user hopes to obtain, with no information about
+the entities that generated the data. They map almost directly into types that
+are used by the underlying Opal system. We add a few types that are noteworthy:
+the type {\bf string} is added to account for FreeText entities; and $\langle
+{\ell} \rangle$ represents a disjunction of literals, and covers the case of
+Keywords entities.
 
@@
 \begin{align*}
-  &r\ \text{well-formed}_{\text{cfg}} \iff \\
+  &r\ \text{well-formed}_{\ff{cfg}} \iff \\
   &\forall (\etag{t} : \textsf{Tag}).\ \ff{cfg}(\etag{t})\ \text{defined} \Rightarrow \ff{lookup}(\etag{t}, r) : \ff{cfg}(\etag{t})
 \end{align*}
+
+When constructing a member of \ff{Cfg}, we must take care that the mapping
+respects entity types. Thus, if \etag{t} is a tag for a free-text entity,
+$\ff{cfg}(\etag{t})$ is allowed to be any string. If, instead, \etag{t} is a tag
+for keywords or trait entities, $\ff{cfg}(\etag{t})$ should return a union of
+literals that represents the appropriate data.
 
@@
 
-We will define \ff{parse} by case analysis on the declaration.
+In the event that $r$ is not well-formed with respect to the configuration
+generated by $d$, $\ff{parse}(d, r)$ is undefined. Otherwise, we define
+\ff{parse} by case analysis on the declaration.
 \begin{align*}
@@
 \subsection{Correctness}
-Given a declaration $d$, we generate a configuration, \ff{cfg}. Then,
+We would like to bring both parts of the formalism together to show that our
+system works the way we would like it to. The main property that we would like
+to show is that \ff{parse} is correct.
+\begin{theorem}[\ff{parse} Correct]
+  Let $d$ be a declaration in $\fcy{D}$, and generate a configuration, \ff{cfg},
+  from $d$. Then,
   \begin{flalign*}
     &\forall\ (r : \textsf{Resp}).\\
-  &r\ \text{well-formed}_{\text{cfg}} \Rightarrow \text{parse}(d, r)\ \text{defined}
+    &r\ \text{well-formed}_{\ff{cfg}} \Rightarrow \ff{parse}(d, r)\ \text{defined}
   \end{flalign*}
+\end{theorem}
+\begin{proof}
+  By induction on $d$.
+\end{proof}
 Note that since \ff{parse} is a dependent function, it will already return a
-value of the correct type. Thus, we only need to care that it is defined. Proof
-of this fact is clear by induction on $d$.
+value of the correct type. Thus, we only need to care that it is defined.
 

### 2017-12-13 paper.tex
Added a sentence to the conclusion
Started working on Adrian's suggestions

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{amsthm}
@@
 \usepackage{subcaption}
+\usepackage{cite}
 
@@
 \end{itemize}
-This is the realm of Natural Language Understanding (NLU)\footnote{We do not
+This is the realm of Natural Language Understanding (NLU).\footnote{We do not
   distinguish between spoken or written language when talking about NLU in
-  general.}.
+  general.}
 
@@
 this power available at the language level, so programmers have been forced to
-integrate these technologies into their applications by hand. (In fact, machine
-learning in general has been relatively neglected by the programming languages
-community.) We propose an addition to the Opal language that will begin to close
-this gap.
+integrate these technologies into their applications by hand. We propose an
+addition to the Opal language that will begin to close this gap.
 
@@
 message and a time (for simplicity, we will just allow the user to specify
-\ts{later} or \ts{tomorrow}) or she
-might ask the application to get any active reminders. Since Opal is embedded
-in Typescript, a typical Opal application might express this interface with the
-following declarations:\footnote{An overview of Typescript's type system can be
-  found at \url{www.typescriptlang.org}.}
+\ts{later} or \ts{tomorrow}) or she might ask the application to get any active
+reminders. Since Opal is embedded in TypeScript,\cite{typescript} a typical Opal
+application might express this interface with the following declarations:
 \begin{minted}{typescript}
@@
 We will begin by presenting a technical description of our type language in
-section \ref{description}. In section \ref{formalism}, we present a formal
+Section \ref{description}. In Section \ref{formalism}, we present a formal
 specification for the language our type-directed response parsing algorithm.
 Section \ref{implementation} covers the implementation of the language and
-related libraries. Finally, sections \ref{evaluation} and \ref{future} present
+related libraries. Finally, Sections \ref{evaluation} and \ref{future} present
 an evaluation of our work, and our plans for future work, respectively.
@@
 single NLU backend---specifically, we are working with an NLU engine called
-Wit.\fxnote{[CITE]} When configuring Wit, as with many NLU backends, the main
-task is setting up the \emph{entities} that Wit should be looking for. An entity
-is a discrete piece of information that the NLU engine extracts from an
-utterance. In our example, the messages and the time are both entites, as is the
-user's intent.
+Wit.\cite{wit} When configuring Wit, as with many NLU backends, the main task is
+setting up the \emph{entities} that Wit should be looking for. An entity is a
+discrete piece of information that the NLU engine extracts from an utterance. In
+our example, the messages and the time are both entites, as is the user's
+intent.
 
 \subsection{Search Strategies}
-When entities are declared, the user must specify a ``search strategy''. In the
+When entities are declared, the user must specify a ``search strategy.'' In the
 case of Wit, there are three main strategies:
@@
 from before in a slightly different way. Figure \ref{fig:dsl_decls} same type
-declarations from section \ref{introduction} written in our DSL.
+declarations from Section \ref{introduction} written in our DSL.
 
@@
 always tagged unions, we can write them in (almost) ML syntax, rather than
-writing out the idiomatic Typescript workaround.
+writing out the idiomatic TypeScript workaround.
 
 Finally, writing specific entity declarations rather than just \ts{type}, makes
-it clear that there is more going on than a direct translation into Typescript.
+it clear that there is more going on than a direct translation into TypeScript.
 In fact, the DSL parser does lot of validation to guarantee that only reasonable
@@
 to be a union, and that union must contain only literals. Even the \ts{alias}
-keyword has some validation built in. Some Typescript types would be problematic
+keyword has some validation built in. Some TypeScript types would be problematic
 if they were allowed---we enforce those conditions in the DSL.
@@
 \begin{figure}
-\begin{minted}{haskell}
-  data Decl = FreeText String
-            | Keywords String [String]
-            | Trait String [(String, Ty)]
-            | Alias String Ty
-            deriving (Show)
-
-  data Ty = Def String
-          | Lit String
-          | Rec [(String, Ty)]
-          deriving (Show)
-\end{minted}
+  \centering
+  \begin{align*}
+    \mathbb{D} ::=&\ \text{FreeText}(\etag{t}) \\
+    |&\ \text{Keywords}(\etag{t},\ [\ell]) \\
+    |&\ \text{Trait}(\etag{t},\ \{s: \mathbb{T}\}) \\
+    |&\ \text{Alias}(\etag{t},\ \mathbb{T}) \\
+    \\
+    \mathbb{T} ::=&\ \text{Def}(\etag{t}) \mid \ell \mid \{s: \mathbb{T}\}
+  \end{align*}
+  $$ \etag{t} \in \ff{Tag} \qquad \ell, s \in \textbf{string} $$
   \caption{The abstract syntax tree for our DSL.}
@@
 backends provide the ability to parse numbers, dates, and other data, these
-types of entities are implementation specific. In section \ref{future}, we
+types of entities are implementation specific. In Section \ref{future}, we
 discuss plans to expand the allowed types in a modular way.
@@
 
-We elected to use a Typescript library called \emph{runtypes} to serialize the
-AST directly into Typescript; this gives us a data structure to guide response
+We elected to use a TypeScript library called \emph{runtypes} to serialize the
+AST directly into TypeScript; this gives us a data structure to guide response
 parsing, and also provides us a way to dynamically type-check the results. We
-discuss the response parsing algorithm in more detail in section
+discuss the response parsing algorithm in more detail in Section
 \ref{implementation}.
@@
 to parse the declarations. The tool serializes three artifacts: a set of
-Typescript type declarations, a runtypes datastructure, and a set of JSON
+TypeScript type declarations, a runtypes datastructure, and a set of JSON
 configuration files for Wit. The programmer simply imports the types and
@@
 In order to reason about the correctness of our translations, we have formalized
-the results from section \ref{description}. We begin by formalizing the type DSL
-as shown in figure \ref{fig:grammar_a}.
+the results from Section \ref{description}. We begin by formalizing the type DSL
+as shown in Figure \ref{fig:grammar_a}.
 \begin{figure}
@@
   \end{subfigure}
-  \caption{Formal grammars.}
+  \caption{The core language.}
 \end{figure}
-There are similarities between these definitions and the Haskell ones in figure
-\ref{fig:ast}, but these have some slightly different goals. First, we remove
-aliases and require that all type definitions are written ``in-line''. Next,
-rather than split the definition by allowing tags within $\fcy{T}$, we also
-require that entitiy definitions are inlined. This means that the data a user
-expects from the NLU backend can be represented by a single tree of type
+There are similarities between these definitions and the ones in Figure
+\ref{fig:ast}, but the core language has some slightly different goals. First,
+we remove aliases and require that all type definitions are written ``in-line.''
+Next, rather than split the definition by allowing tags within $\fcy{T}$, we
+also require that entity definitions are inlined. This means that the data a
+user expects from the NLU backend can be represented by a single tree of type
 $\fcy{D}$. Importantly, none of these simplifications are fundamental---this
@@
 
-In addition to a grammar for the DSL itself, in figure \ref{fig:grammar_b} we
-define a grammar of types, $\tau$. We can think of $\tau$ as representing the
-final form of the data that a user hopes to obtain, with no information about
-the entities that generated the data. They map almost directly into types that
-are used by the underlying Opal system. We add a few types that are noteworthy:
-the type {\bf string} is added to account for FreeText entities; and $\langle
-{\ell} \rangle$ represents a disjunction of literals, and covers the case of
-Keywords entities.
+In addition to a grammar for the core language itself, in Figure
+\ref{fig:grammar_b} we define a grammar of types, $\tau$. We can think of $\tau$
+as representing the final form of the data that a user hopes to obtain, with no
+information about the entities that generated the data. They map almost directly
+into types that are used by the underlying Opal system. We add a few types that
+are noteworthy: the type {\bf string} is added to account for FreeText entities;
+and $\langle {\ell} \rangle$ represents a disjunction of literals, and covers
+the case of Keywords entities.
 
@@
 Note that the function is partial; there is no guarantee that the response will
-contain the desired data. In the sequel, we will often assume that \ff{lookup}
-is defined on tags we care about, but that assumption will be stated explicitly
-rather than assumed by default.
+contain the desired data. In the following development, we will often assume
+that \ff{lookup} is defined on tags we care about, but that assumption will be
+stated explicitly rather than assumed by default.
 
@@
 fairly automatic. If the type is another declaration, we can call back to
-\ff{parse}, if it is a literal we simply generate the appropriate string, and if
+\ff{parse}: if it is a literal we simply generate the appropriate string; and if
 it is a record, we build the appropriate record and recursively build out its
@@
 library like Wit, they begin by using our configuration tool. The tool turns
-code in the type DSL into Typescript code and JSON configuration, and ensures
+code in the type DSL into TypeScript code and JSON configuration, and ensures
 that the NLU engine responses will agree with the type definitions in the user's
@@
 obviously correct. We use the Parsec library to parse the type DSL into the
-intermediate representation mentioned above in figure \ref{fig:ast}.
+intermediate representation mentioned in Figure \ref{fig:hs_ast}.
+
+\begin{figure}
+\begin{minted}{haskell}
+  data Decl = FreeText String
+            | Keywords String [String]
+            | Trait String [(String, Ty)]
+            | Alias String Ty
+            deriving (Show)
+
+  data Ty = Def String
+          | Lit String
+          | Rec [(String, Ty)]
+          deriving (Show)
+\end{minted}
+  \caption{The Haskell code for our AST.}
+  \label{fig:hs_ast}
+\end{figure}
 
@@
 runtypes object to check the parsed object, and we cast it to the desired type.
-The definition is shown in figure \ref{fig:parse}. The cast is safe, so long as
+The definition is shown in Figure \ref{fig:parse}. The cast is safe, so long as
 the user respects the invariant that parse is always called with compatible type
@@
 information about the matched entities. Wit, in particular, provides confidence
-levels for each entitiy that it finds. Thus, a value like \ts{"Set"} might be
+levels for each entity that it finds. Thus, a value like \ts{"Set"} might be
 paired with a confidence level of $70\%$, meaning that Wit is $70\%$ sure that
@@
 more popular as a user interface, it makes sense that Opal would provide a clean
-and safe interface for natural langauge interaction.
+and safe interface for natural langauge interaction. This work contributes a
+real, working tool in the Opal ecosystem that provides a first attempt at such
+an interface.
+
+\bibliography{paper}{}
+\bibliographystyle{plain}
 

### 2017-12-13 paper.tex
More work on various sections

@@
 \subsection{Types for NLU}
-If one were to try to implement our reminder interface na\"ively, a problem
-would arise: the popular backends provide untyped interfaces and require
-external configuration. Since Opal hopes to provide a consistent and safe
-interface to multiple NLU engines, we decided to develop a typed abstraction for
-natural language systems.
+If a programmer simply wrote down types like the ones above, set up their NLU
+engine, and started their application, there would be a problem. Since there is
+no connection between the types and the NLU configuration, these types could
+actually be unsound! If the programmer made a mistake in \emph{either} the types
+\emph{or} the configuration, the application would be unusable and no compiler
+could possibly know. Bugs like these are particularly insidious because they
+might be mistaken as network errors or even just as an indicator of
+under-training.
 
-In this work, we present a domain specific language (DSL) for simultaneously
+In this work, we set out to connect the types and the NLU configuration,
+allowing the programmer to avoid these awful bugs. There are a few ways that one
+might try to do this. First, the programmer could write down the NLU
+configuration and extract the types from it. This is pretty much a non-starter,
+since there is fundamentally less information in the NLU configuration than
+there is in the type definitions. Instead, the programmer could write the types,
+and extract the NLU information from those. This approach is more feasible, but
+it would require some kinds of annotations in the types to capture
+implementation-specific properties of the configuration.
+
+We present a third option: a domain specific language (DSL) for simultaneously
 specifying
@@
 \end{itemize}
-We also implement translations from our DSL to various artifacts which can be
-integrated with Opal applications.
+This allows the user to tune the configuration as much as they need, to write
+useful types, and to ensure that these two artifacts are always compatible.
 \\
@@
 \begin{minted}{typescript}
-  free-text Msg;
-  keywords Time = "later" | "tomorrow";
+  free-text Msg
+  keywords Time = "later" | "tomorrow"
   alias Set = {
@@
     time: Time
-  };
+  }
   trait Intent =
     | <Set> Set
-    | <Get> {};
+    | <Get> {}
 \end{minted}
@@
   {
-    Intent: "Set",
-    Msg: "pick up the dry cleaning",
-    Time: "later"
+    Intent: ["Set"],
+    Msg: ["pick up the dry cleaning"],
+    Time: ["later"]
   }
@@
 
-In order to reconstruct the desired data from the NLU backend's response, we
-realized that it was necessary to keep data from the DSL around at run-time, not
-just during setup.
+This makes it necessary to have a run-time component of the system that knows
+how to reconstruct structured data from the unstructured response that comes
+from the NLU engine. Building structured data from the response is a nontrivial
+task. First, it requires that information about the relationships between
+entities and data be available at run-time. This is mostly an engineering
+problem, and our solution is outlined in Section \ref{implementation}.
 
-\begin{figure*}
-  \centering
-  \begin{tikzpicture}[node distance = 3cm, auto]
-    \node [block] (types) {Type DSL};
-    \node [block, right of=types] (decls_hs) {Decl (hs)};
-    \node [block, right of=decls_hs] (decls_ts) {Runtypes (ts)};
-    \node [block, below of=decls_ts] (cfg) {NLU Config};
-    \node [block, above of=decls_ts] (ts_types) {Types (ts)};
-    \node [cloud, right of=decls_ts] (ts) {Opal};
-    \node [cloud, right of=cfg] (backend) {Wit};
-
-    \path [line] (types) -- (decls_hs);
-    \path [line] (decls_hs) -- (decls_ts);
-    \path [line] (decls_hs) -- (cfg);
-    \path [line] (decls_hs) -- (ts_types);
-    \path [line] (decls_ts) -- (ts);
-    \path [line] (ts_types) -- (ts);
-    \path [line] (cfg) -- (backend);
-  \end{tikzpicture}
-  \caption{The configuration process.}
-  \label{fig:process}
-\end{figure*}
-
-We elected to use a TypeScript library called \emph{runtypes} to serialize the
-AST directly into TypeScript; this gives us a data structure to guide response
-parsing, and also provides us a way to dynamically type-check the results. We
-discuss the response parsing algorithm in more detail in Section
-\ref{implementation}.
-
-Figure \ref{fig:process} shows process of configuring an application for NLU
-interaction. The programmer writes types in our DSL, and uses our Haskell tool
-to parse the declarations. The tool serializes three artifacts: a set of
-TypeScript type declarations, a runtypes datastructure, and a set of JSON
-configuration files for Wit. The programmer simply imports the types and
-runtypes into their Opal project, uploads the JSON files to Wit, and the system
-is ready to go.
+The other difficulty is that the types might admit more than one valid response
+object. To see how this works, imagine that we want to allow a user to set two
+messages at once. We would change the \ts{Set} type to be:
+\begin{minted}{typescript}
+  type Set = {
+    msg1: Msg,
+    msg2: Msg,
+    time: Time
+  };
+\end{minted}
+Unfortunately, this kind of ambiguity makes it impossible to find a principled
+way of parsing a response. Since the backends do not expose ordering
+information, a response with two \ts{Msg} fields might parse to any of four
+different objects. We say four because the reconstruction algorithm might
+reasonably ignore one of the messages and just put the same one for \ts{msg1}
+and \ts{msg2}. For this reason, we impose the restriction that declarations do
+not reuse entities.
 
@@
   \begin{subfigure}{1\linewidth}
-    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \ell
+    $$ \tau ::=\ \textbf{string} \mid \ell \mid \{s: \tau\} \mid \langle \tau
     \rangle $$
@@
 are noteworthy: the type {\bf string} is added to account for FreeText entities;
-and $\langle {\ell} \rangle$ represents a disjunction of literals, and covers
-the case of Keywords entities.
+and $\langle {\tau} \rangle$ represents an arbitrary union, and covers the cases
+of Keywords and Trait entities.
 
@@
 \end{align*}
+We implement \ff{interp} and \ff{interpTy} in Figure \ref{fig:interp}.
+
+\begin{figure}
+  \centering
+  \begin{flalign*}
+    & \ff{interp}(\text{FreeText}(\etag{t})) = \textbf{string} \\
+    & \ff{interp}(\text{Keywords}(\etag{t}, [\ell])) = \langle \ell \rangle \\
+    & \ff{interp}(\text{Trait}(\etag{t'}, \{s : t\})) = \\
+    & \qquad \left< \{\text{tag}: s,\ \text{data}: \ff{interpTy(t)}\} \right> \\
+    \\
+    & \ff{interpTy}(d) = \ff{interp(d)} \\
+    & \ff{interpTy}(\ell) = \ell \\
+    & \ff{interpTy}(\{s: t\}) = \{s: \ff{interpTy}(t)\}
+  \end{flalign*}
+  \caption{Type interpretation.}
+  \label{fig:interp}
+\end{figure}
+
+One interesting thing to point out is that the record in the Trait declarations
+becomes a union type. This makes sense, since a trait represents a property of
+the utterance as a whole, and thus can decide the structure of the response.
 
 With the basic building blocks defined, we can move forward in formalizing our
-system. \fxnote{Should I write out implementations?}
+system.
 
@@
 We define configurations as functions that map a subset of entity tags to the
-type of data expected for those entities.
-$$ \ff{Cfg} \equiv \ff{Tag} \rightharpoonup \tau $$
-Now, we call a response well-formed with respect to $\ff{cfg} : \ff{Cfg}$ if and
-only if it returns appropriate data on \ff{lookup}. More formally, for some $r :
+type of data expected for those entities. The function \ff{cfg} takes a
+declaration and produces a configuration for that declaration.
+$$ \ff{cfg}\ :\ \fcy{D} \to \ff{Tag} \rightharpoonup \tau $$
+We define \ff{cfg} in Figure \ref{fig:cfg}. The exact details of the else branch
+of the Trait case are implementation specific, but the basic idea to map the
+\ff{cfg} function over all nested declarations. This achieves the goal of
+capturing every tag in the declaration.
+
+\begin{figure}
+  \centering
+  \begin{flalign*}
+    & \ff{cfg}(\text{FreeText}(\etag{t}), \etag{t}) = \textbf{string} \\
+    & \ff{cfg}(\text{Keywords}(\etag{t}, [\ell]), \etag{t}) = \langle \ell \rangle \\
+    & \ff{cfg}(\text{Trait}(\etag{t}', \{s : t\}), \etag{t}) = \\
+    & \qquad \textbf{if}\ \etag{t} = \etag{t}'\ \textbf{then}\ \langle {s} \rangle\ \textbf{else}\ \dots
+  \end{flalign*}
+  \caption{Configuation generation.}
+  \label{fig:cfg}
+\end{figure}
+
+Now, we call a response well-formed with respect to $d : \fcy{D}$ if and only if
+it returns appropriate data on \ff{lookup}. More formally, for some $r :
 \ff{Resp}$,
 \begin{align*}
-  &r\ \text{well-formed}_{\ff{cfg}} \iff \\
-  &\forall (\etag{t} : \textsf{Tag}).\ \ff{cfg}(\etag{t})\ \text{defined} \Rightarrow \ff{lookup}(\etag{t}, r) : \ff{cfg}(\etag{t})
+  &r\ \text{well-formed}_{d} \iff \\
+  &\ \ \forall (\etag{t} : \textsf{Tag}).\ \ff{cfg}(d, \etag{t})\ \text{defined} \Rightarrow  \\
+  &\qquad \ff{lookup}(\etag{t}, r) : \ff{cfg}(d, \etag{t})
 \end{align*}
-
-When constructing a member of \ff{Cfg}, we must take care that the mapping
-respects entity types. Thus, if \etag{t} is a tag for a free-text entity,
-$\ff{cfg}(\etag{t})$ is allowed to be any string. If, instead, \etag{t} is a tag
-for keywords or trait entities, $\ff{cfg}(\etag{t})$ should return a union of
-literals that represents the appropriate data.
+We will use well-formedness as our basic assumption that the application has
+been configured correctly relative to the response.
 
@@
 The final case for \ff{parse} simply matches the value in the response to an
-index into the record, and interprets that record entry using \ff{parseTy}.
+index into the record, and interprets that record entry using \ff{parseTy}. It
+is important to note that since \ff{interp} returns a union here, we only need
+to return a value that is in some branch of that union. In this case, we use the
+\ff{lookup} value to decide which branch.
 
@@
 \begin{theorem}[\ff{parse} Correct]
-  Let $d$ be a declaration in $\fcy{D}$, and generate a configuration, \ff{cfg},
-  from $d$. Then,
+  Let $d$ be a declaration in $\fcy{D}$. Then,
   \begin{flalign*}
     &\forall\ (r : \textsf{Resp}).\\
-    &r\ \text{well-formed}_{\ff{cfg}} \Rightarrow \ff{parse}(d, r)\ \text{defined}
+    &r\ \text{well-formed}_d \Rightarrow \ff{parse}(d, r)\ \text{defined}
   \end{flalign*}
@@
 \section{Implementation} \label{implementation}
+\begin{figure*}
+  \centering
+  \begin{tikzpicture}[node distance = 3cm, auto]
+    \node [block] (types) {Type DSL};
+    \node [block, right of=types] (decls_hs) {Decl (hs)};
+    \node [block, right of=decls_hs] (decls_ts) {Runtypes (ts)};
+    \node [block, below of=decls_ts] (cfg) {NLU Config};
+    \node [block, above of=decls_ts] (ts_types) {Types (ts)};
+    \node [cloud, right of=decls_ts] (ts) {Opal};
+    \node [cloud, right of=cfg] (backend) {Wit};
+
+    \path [line] (types) -- (decls_hs);
+    \path [line] (decls_hs) -- (decls_ts);
+    \path [line] (decls_hs) -- (cfg);
+    \path [line] (decls_hs) -- (ts_types);
+    \path [line] (decls_ts) -- (ts);
+    \path [line] (ts_types) -- (ts);
+    \path [line] (cfg) -- (backend);
+  \end{tikzpicture}
+  \caption{The configuration process.}
+  \label{fig:process}
+\end{figure*}
+A high-level workflow for our implementation is shown in Figure
+\ref{fig:process}. The programmer writes types in our DSL, and uses our Haskell
+tool to parse the declarations. The tool serializes three artifacts: a set of
+TypeScript type declarations, a runtypes datastructure, and a set of JSON
+configuration files for Wit. The programmer simply imports the types and
+runtypes into their Opal project, uploads the JSON files to Wit, and the system
+is ready to go.
 
@@
 obviously correct. We use the Parsec library to parse the type DSL into the
-intermediate representation mentioned in Figure \ref{fig:hs_ast}.
+intermediate representation mentioned in Figure \ref{fig:hs_ast}.\cite{parsec}
 
@@
   data Decl = FreeText String
-            | Keywords String [String]
-            | Trait String [(String, Ty)]
+            | Keywords String
+                       [String]
+            | Trait String
+                    [(String, Ty)]
             | Alias String Ty
@@
 \end{figure}
+
+After parsing with Parsec, we apply versions of the \ff{interp} and \ff{cfg}
+functions from Section \ref{formalism} to generate the appropriate
+configuration.
 
@@
 of $\fcy{D}$ is the \hs{Decl} type in the Haskell code; this is not available to
-the Opal application at run time. As mentioned above, we get around this by
-using a modified version the runtypes library to keep this information around at
-run-time. We extend runtypes to include a type for entities---which map to
+the Opal application at run time.
+
+We get around this by using a modified version the a TypeScript library called
+Runtypes.\cite{runtypes} Runtypes allows us to to serialize declarations
+directly into TypeScript and therefore to keep necessary information around at
+run-time. We extend Runtypes to include a type for entities---which map to
 declarations in $\fcy{D}$. Runtypes provides the infrastructure for non-entity
@@
 \section{Evaluation} \label{evaluation}
-%TODO: Cover limitations of the design, possible use-cases, etc
+Working with the system to build some toy applications led us to a number of
+important conclusions about its effectiveness. We discuss a number of these
+conclusions here.
+
+\subsection{Type Expressiveness}
+The first important thing we found when starting to write types in our DSL is
+that the standard three entity types are not quite expressive enough for real
+applications. For example, when trying to implement a small chatbot for
+scheduling calendar meetings, we wanted to write something like,
+\begin{minted}{typescript}
+  free-text Person
+  ...
+  trait Intent = <Schedule> {
+    person: Person,
+    date: Date,
+    time: Time
+  }
+\end{minted}
+but we found that we had no way to express a date or a time. Technically
+speaking, it might have been possible to encode dates and times as free-text and
+do a second parsing pass in the application code, but that solution is far from
+elegant. We discuss one potential solution to this problem in Section
+\ref{future>types}.
+
+\subsection{Programmer Freedom}
+When working around unsupported types, there is an obvious tendency to modify
+the application types to better fit the capabilities of the system. It turns out
+that this occurs in more cases than just using dates and times. For example,
+consider another calendar example, similar to the one above.
+\begin{minted}{typescript}
+keywords WkDay = "Sunday" | ...
+alias Abs = {
+  day: WkDay
+  person: Person
+}
+keywords RDay = "today" | "tomorrow"
+alias Rel = {
+  day: RDay
+  person: Person
+}
+trait Intent =
+  | <Absolute> Abs
+  | <Relative> Rel
+\end{minted}
+Here, we've done away with the complications of dates and times, and we just
+allow a user to specify a day of the week or a ``relative day.'' The interesting
+observation here is that this is likely not the way that most TypeScript/Opal
+developers would write this type. For example, one reasonable alternative might
+express days as a top-level disjunction, and then make the type of intents a
+normal record.
+\begin{minted}{typescript}
+  type Day =
+    | { tag: "today" }
+    | { tag: "tomorrow" }
+    | { tag: "abs", day: WkDay }
+  type Intent =
+    { day: Day, person: Person }
+\end{minted}
+This makes sense, since \ts{Absolute} and \ts{Relative} are not actully
+different \emph{intents}, per se. We can see that the system has led us to a
+different type than we would normally write.
+
+It turns out that this might actually be a good thing, if viewed from the
+perspective of the NLU engine itself. While \ts{Absolute} and \ts{Relative} are
+the same intent from a philosophical perspective, they might be parsed very
+differently. The utterance
+\begin{center}
+  \emph{Schedule my Wednesday meeting with Chris.}
+\end{center}
+is very different in structure from
+\begin{center}
+  \emph{Set a meeting with James tomorrow.}
+\end{center}
+Thus, it might be the case that forcing the programmer to write slightly
+different types will make it much easier for the NLU backend to find the correct
+information.
+
+Overall, there is likely some work to be done in terms of making our types more
+idiomatic, but for now we err on the side of making life easier for the NLU
+engine.
 
@@
 
-\subsubsection{More Types}
+\subsubsection{More Types} \label{future>types}
 In addition to the three ``universal'' kinds of entities (free-text, keywords,
@@
 
+% TODO: Fix bib formatting
 \bibliography{paper}{}

### 2017-12-13 paper.tex
Edits/revisions

@@
 today never touch a command shell, and even those who do would often rather not.
-Graphical interfaces are amazingly powerful, and are often good enough to allow
+Graphical interfaces are amazingly powerful, and are often sufficient to allow
 users to easily interact with the computer, but even those have their drawbacks.
@@
 \subsection{An Example}
-Consider a simple reminder interface to a calendar or other scheduling
+Consider a simple \emph{reminders} interface to a calendar or other scheduling
 assistant. A user might ask the application to set a reminder containing a
@@
 
-Since we would like to interact with this application via human language, we
-need some way of converting a user \emph{utterance} to something of type
-\ts{Intent}. Figure \ref{fig:drycleaning} shows the desired
-data representation of a user utterance.
+We would like to interact with this application via human language, so we need
+some way of converting a user's \emph{utterance} to something of type
+\ts{Intent}. Figure \ref{fig:drycleaning} shows the desired data representation
+of a user utterance.
 
@@
   a quality of the utterance as a whole. Something like a user's intent or even
-  mood could be a trait.
+  their mood could be a trait.
 \end{description}
@@
 With our observations about search strategies in mind, we can write our types
-from before in a slightly different way. Figure \ref{fig:dsl_decls} same type
-declarations from Section \ref{introduction} written in our DSL.
+from before in a slightly different way. Figure \ref{fig:dsl_decls} shows the
+same type declarations from Section \ref{introduction} written in our DSL.
 
@@
 The main thing to notice is that the keyword \ts{type} has been replaced by
-entity declarations \ts{free-text}, \ts{keywords}, and \ts{trait}, as well as
+entity declarations \ts{free-text}, \ts{keywords}, and \ts{trait}, or by
 \ts{alias}. There are a number of benefits of entity declarations. First and
@@
 Another benefit of these bespoke declaration forms is simplified syntax. When a
-\ts{free-text} entity is declared, there is no need to write ``\ts{= string}'',
+\ts{free-text} entity is declared, there is no need to write ``\ts{= string},''
 because that is implied by the search strategy. Similarly, since traits are
 always tagged unions, we can write them in (almost) ML syntax, rather than
-writing out the idiomatic TypeScript workaround.
+writing out the standard TypeScript workaround.
 
@@
 in a trait declaration are fairly strict. As of writing, there are only three
-types allowed within a trait declaration: previously defined entities, literals,
-and records (where each element is subject to the same restrictions). These
-limitations match up with the capabilities of NLU engines in general. While many
-backends provide the ability to parse numbers, dates, and other data, these
-types of entities are implementation specific. In Section \ref{future}, we
+types allowed within a trait declaration: previously defined declarations,
+literals, and records (where each element is subject to the same restrictions).
+These limitations match up with the capabilities of NLU engines in general.
+While many backends provide the ability to parse numbers, dates, and other data,
+these types of entities are implementation specific. In Section \ref{future}, we
 discuss plans to expand the allowed types in a modular way.
@@
 In order to reason about the correctness of our translations, we have formalized
-the results from Section \ref{description}. We begin by formalizing the type DSL
-as shown in Figure \ref{fig:grammar_a}.
+the results from Section \ref{description}. We begin by writing down a core
+language for the type DSL as shown in Figure \ref{fig:grammar_a}.
 \begin{figure}
@@
     $$ \etag{t} \in \ff{Tag} \qquad \ell, s \in \textbf{string} $$
-    \caption{Type DSL}
+    \caption{The core language.}
     \label{fig:grammar_a}
@@
     \rangle $$
-    \caption{Relevant Opal Types}
+    \caption{Opal types.}
     \label{fig:grammar_b}
   \end{subfigure}
-  \caption{The core language.}
+  \caption{Formal structures.}
 \end{figure}
@@
 into types that are used by the underlying Opal system. We add a few types that
-are noteworthy: the type {\bf string} is added to account for FreeText entities;
-and $\langle {\tau} \rangle$ represents an arbitrary union, and covers the cases
-of Keywords and Trait entities.
+are noteworthy. The type {\bf string} is added to account for FreeText entities.
+The type $\langle {\tau} \rangle$ represents an arbitrary union and covers the
+cases of Keywords and Trait entities.\footnote{This notation is slightly
+unconventional, but we prefer to write $\langle \tau \rangle$ or $\langle \tau_1
+\mid \dots \mid \tau_n \rangle$ rather than simply $\tau_1 \mid \dots \mid
+\tau_n$ to separate it from the surrounding syntax.}
 
@@
 becomes a union type. This makes sense, since a trait represents a property of
-the utterance as a whole, and thus can decide the structure of the response.
+the utterance as a whole, and thus can decide the structure of the rest of the
+response.
 
@@
 number of things. The data might be a particular literal, in the case of a
-keywords or trait entity, or an entire phrase in the case of a free-text entity.
+keywords or trait entity, or an entire phrase in the case of a free text entity.
 We define configurations as functions that map a subset of entity tags to the
@@
 Now, we call a response well-formed with respect to $d : \fcy{D}$ if and only if
-it returns appropriate data on \ff{lookup}. More formally, for some $r :
-\ff{Resp}$,
+it returns appropriate data on \ff{lookup}. More formally, we say that for some
+$r : \ff{Resp}$,
 \begin{align*}
@@
 \end{align*}
-We will use well-formedness as our basic assumption that the application has
-been configured correctly relative to the response.
+We will use well-formedness as our basic signal that a the application and the
+response have been configured to be compatible.
 
@@
 \end{align*}
-This case is fairly straightforward. Since the interpretation of a free-text
+This case is fairly straightforward. Since the interpretation of a free text
 entity is just a {\bf string}, we can simply look up the value of the
@@
     \node [block, right of=types] (decls_hs) {Decl (hs)};
-    \node [block, right of=decls_hs] (decls_ts) {Runtypes (ts)};
+    \node [block, right of=decls_hs] (decls_ts) {AST (ts)};
     \node [block, below of=decls_ts] (cfg) {NLU Config};
@@
 A high-level workflow for our implementation is shown in Figure
-\ref{fig:process}. The programmer writes types in our DSL, and uses our Haskell
+\ref{fig:process}. The programmer writes types in our DSL and uses our Haskell
 tool to parse the declarations. The tool serializes three artifacts: a set of
-TypeScript type declarations, a runtypes datastructure, and a set of JSON
-configuration files for Wit. The programmer simply imports the types and
-runtypes into their Opal project, uploads the JSON files to Wit, and the system
-is ready to go.
+TypeScript type declarations, an AST datastructure, and a set of JSON
+configuration files for Wit. The programmer simply imports the types and AST
+into their Opal project, uploads the JSON files to Wit, and the system is ready
+to go.
 
@@
 When a user wants to configure an Opal application to interact with an NLU
-library like Wit, they begin by using our configuration tool. The tool turns
-code in the type DSL into TypeScript code and JSON configuration, and ensures
+library like Wit, they begin by using our configuration tool. The tool ensures
 that the NLU engine responses will agree with the type definitions in the user's
@@
 
-\subsection{Runtime Library}
-At run-time typical application using NLU will go through a fairly consistent
+\subsection{Run-time Library}
+At run-time, a typical application using NLU will go through a fairly consistent
 cycle.
@@
 network code involved in obtaining the response is fairly uninteresting, so we
-will focus on the parsing,
+will focus on the parsing.
 
@@
 The other difference arises because Opal does not have dependent types. We
-cannot completely faithfully transcribe the parse function from above, because
-it is not obvious that it can be typed at all. The solution here is to combine
-Opal's dynamic typing with runtypes' dynamic checking feature to create a
-function that appears to be dependently typed. First, we use dynamic types
-within the parse function. While building the response data, we assume, where
+cannot completely faithfully transcribe the \ff{parse} function from above,
+because it is not obvious that it can be typed at all. The solution here is to
+combine Opal's dynamic typing with Runtypes' dynamic checking feature to create
+a function that appears to be dependently typed. First, we use dynamic types
+within the \ff{parse}. While building a response object, we assume, where
 necessary, that it has type \ts{any}. Then, once parsing is complete, we use the
-runtypes object to check the parsed object, and we cast it to the desired type.
-The definition is shown in Figure \ref{fig:parse}. The cast is safe, so long as
-the user respects the invariant that parse is always called with compatible type
-and runtype parameters.
+Runtypes object to check the parsed result, and we cast the result to the
+desired type. The definition is shown in Figure \ref{fig:parse}. The cast is
+safe, so long as the user respects the invariant that parse is always called
+with compatible type and runtype parameters.
 
@@
 but we found that we had no way to express a date or a time. Technically
-speaking, it might have been possible to encode dates and times as free-text and
+speaking, it might have been possible to encode dates and times as free text and
 do a second parsing pass in the application code, but that solution is far from
@@
 different types will make it much easier for the NLU backend to find the correct
-information.
+information. There is likely some work to be done in terms of making our types
+more idiomatic, but for now we err on the side of making life easier for the NLU
+engine.
 
-Overall, there is likely some work to be done in terms of making our types more
-idiomatic, but for now we err on the side of making life easier for the NLU
-engine.
+\subsection{Relative Success}
+There is still a lot to be done on this work and in this space, but as of
+writing, we feel that we have proven the concept fairly well. The system does
+the job it intends to, and is usable for actual applications. In general, using
+the tools feels fairly natural. The Haskell tool produces the output that a user
+would expect, and the programming interface provides the functionality that a
+user needs to actually write an application that uses NLU. Overall, we are happy
+with our approach to the stated problem as well as with the results of the
+implementation.
 
@@
 \subsubsection{More Types} \label{future>types}
-In addition to the three ``universal'' kinds of entities (free-text, keywords,
+In addition to the three ``universal'' kinds of entities (free text, keywords,
 and traits), the Wit NLU engine provides more specialized ways of looking up

### 2017-12-13 paper.tex
Updated bib
Moved a figure, small revisions

@@
 integrate these technologies into their applications by hand. We propose an
-addition to the Opal language that will begin to close this gap.
+addition to the Opal language that will begin to close this
+gap.\cite{opal-arxiv}
 
@@
 
-\begin{figure}
+\begin{figure}[H]
 \begin{minted}{haskell}
@@
 that this occurs in more cases than just using dates and times. For example,
-consider another calendar example, similar to the one above.
+consider a different set of types for calendar scheduling.
 \begin{minted}{typescript}
@@
 observation here is that this is likely not the way that most TypeScript/Opal
-developers would write this type. For example, one reasonable alternative might
+developers would write that type. For example, one reasonable alternative might
 express days as a top-level disjunction, and then make the type of intents a
@@
 
-% TODO: Fix bib formatting
 \bibliography{paper}{}

### 2017-12-13 paper.tex
Added a paragraph to the conclusion
(It's a little dramatic)

@@
 
+As we move towards a future where the majority of our interactions with
+computers happen via human language, projects like this will become more and
+more important. It may be that the approach put forth in the paper paves the way
+for safe, correct applications that work with NLU. However, even if a different
+approach wins out, it is almost certain that safe interfaces to NLU engines will
+eventually become wide-spread. Just as companies have tended to favor strongly
+typed language for applications where correctness matters, programmers working
+with NLU will eventually see the need to do so in a principled way.
+
 \bibliography{paper}{}

### 2017-12-14 paper.tex
Changed names to Alice and Bob
Like a good CS researcher
Added a paragraph on hypothetical worlds

@@
 \begin{center}
-  \emph{Schedule my Wednesday meeting with Chris.}
+  \emph{Schedule my Wednesday meeting with Alice.}
 \end{center}
@@
 \begin{center}
-  \emph{Set a meeting with James tomorrow.}
+  \emph{Set a meeting with Bob tomorrow.}
 \end{center}
@@
 
+\noindent \emph{Hypothetical Worlds}\qquad It turns out that this process of
+choosing the parse with the best confidence score is a great application of one
+of Opal's major features---hypothetical worlds. In a perfect world, we would
+allow our parser to generate all possible response objects, compute some
+function based on the confidence values for the entities in the objecs, and
+choose the object with the best value of the function. Writing this code out by
+hand would be tedious, inefficient, and error-prone. Hypothetical worlds give
+this semantics as a language-level feature. When code is executed in a \ts{hyp}
+block, its computation is entirely hypothetical. Multiple \ts{hyp} blocks can be
+run in parallel, and their computations can be observed at specified points.
+Once a particular world is chosen (in this case, by observing the confidence
+function), a sigle world can be ``committed.'' From that point forward, only the
+committed world is treated as real, and the computation can continue with the
+chosen response object.
+
+In the future, we think that hypothetical worlds will open up new opportunities
+for NLU interaction, including the ability to parse responses into more complex
+types.
+
 \subsubsection{Training with Tests}
